\documentclass[11pt,a4paper,parskip=half-]{scrartcl}
\input{../preamble}

\title{The \sire\ language syntax}
\author{James Hanlon}
\date{Last revised \today}

\begin{document}

\section{Collected syntax}

The following sections list the \sire\ syntax, with each section related to a
particular feature.

\input{collected.tex}


\clearpage
\section{Ordered syntax}

The following lists each element of the \sire\ syntax in alphabetical order.

\input{ordered.tex}


\clearpage
\section{Operators}
\label{sec:operators}

A $\pn{binary-operator}$ $\oplus$ takes two operands $a$ and $b$ and produces a
value $a \oplus b$.
%
A binary \emph{arithmetic operators} takes signed integer operands and produces
a signed integer result.

\begin{center}
\begin{tabular}{ccc}
{\bf Symbol} & {\bf Meaning} & {\bf Definition}\\
\midrule
\ttt{+} & sum & $a+b$\\
\ttt{-} & difference & $a-b$\\ 
\ttt{*} & produce & $a\times b$\\
\ttt{/} & quotient & $a / b$\\
\ttt{rem} & remainder & $a \mod b$
\end{tabular}
\end{center}

\noindent 
A binary \emph{relational operator} takes two signed integer values
and produces the value \ttt{true} or \ttt{false}. 

\begin{center}
\begin{tabular}{ccc}
{\bf Symbol} & {\bf Meaning} & {\bf Definition}\\
\midrule
\ttt{=} & equality & $a=b$\\
\ttt{\textasciitilde=} & inequality & $a\neq b$\\
\ttt{<} & less than & $a < b$\\
\ttt{<=} & less than or equal & $a \leq b$\\
\ttt{>} & greater than & $a > b$\\
\ttt{>=} & greater than or equal & $a \geq b$
\end{tabular}
\end{center}

\noindent
A binary \emph{logical operator} takes two operands and produces a bitwise
result $b_i = a_i\oplus b_i$ for $0\leq i < n$.

\begin{center}
\begin{tabular}{ccc}
{\bf Symbol} & {\bf Meaning} & {\bf Definition}\\
\midrule
\ttt{or} & bitwise or & $b$ \ttt{or} $0$ $=$ $b$, $b$ \ttt{or} $1$ $=$ $1$\\
\ttt{and} & bitwise and & $b$ \ttt{and} $0$ $=$ $0$, $b$ \ttt{and} $1$ $=$ $b$\\
\ttt{xor} & bitwise exclusive or & $b$ \ttt{xor} $0$ $=$ $b$, 
  $b$ \ttt{xor} $1$ $=$ $1$, $1$ \ttt{xor} $1$ $=$ $0$, \\
\ttt{{<}<} & left bitwise shift & $b_i=b_{i+1}$, $b_0=0$\\
\ttt{>{>}} & right bitwise shift & $b_i=b_{i-1}$, $b_n-1=0$\\
\end{tabular}
\end{center}

\noindent
A $\pn{unary-operator}$ $\oplus$ takes a single operand $a$ and produces the value
$\oplus a$.

\begin{center}
\begin{tabular}{ccc}
{\bf Symbol} & {\bf Meaning} & {\bf Definition}\\
\midrule
\ttt{-} & negation & $0-a$ \\
\ttt{\textasciitilde} & bitwise not & 
    $\text{\ttt{\textasciitilde}}0=1$, $\text{\ttt{\textasciitilde}}1=0$\\
\end{tabular}
\end{center}


\section{Representation of values}
\label{appx:representation-values}

Signed values are represented as a two's complement bit-pattern. With a word
size of $n$, values in the range $-2^{n-1} \leq n < 2^{n-1}$ can be
represented.

The literal value \ttt{true} represents an all-1 bit pattern (the value $-1$) and the
literal value \ttt{false} represents an all-0 bit pattern (the value 0). The effect of this
is consistent with the logical not operation: \ttt{not} \ttt{false} $=$
\ttt{true} and \ttt{not} \ttt{true} = \ttt{false}.


\section{Character set}

A $\pn{character}$ can be any alphabetical character

\begin{flalign*}
  \hspace{5mm}&\sm{A B C D E F G H I J K L M N O P Q R S T U V W X Y Z}&\\
  \hspace{5mm}&\sm{a b c d e f g h i j k l m n o p q r s t u v w x y z}&
\end{flalign*}

\noindent
or any special character

\begin{flalign*}
  \hspace{5mm}&\sm{\_ + - = , . : ; ? \{ \} [ ] ( ) \# \& ! * @ | " '}&
\end{flalign*}

\noindent
A $\pn{digit}$ can be any numeric character

\begin{flalign*}
  \hspace{5mm}&\sm{0 1 2 3 4 5 6 7 8 9}&
\end{flalign*}

\noindent
A $\pn{hex-digit}$ can be any numeric character or

\begin{flalign*}
  \hspace{5mm}&\sm{A B C D E F a b c d e f}&
\end{flalign*}

\noindent
A $\pn{name}$ consists of a sequence of alphanumeric characters and underscores.


\section{Comments}

A comment is introduced with a `\ttt{\%}' symbol and continues until the end of
  the line.


\begin{minipage}{\textwidth} % keep with next
\section{Keywords}

The following are keywords in \sire\ and cannot be used for names.

\input{keywords}

\end{minipage}

\end{document}
